---
title: "Final Project"
subtitle: "INFO 526 - Summer 2026 - Final Project"
author: 
  - name: "Zachary Zhicheng Xu"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "US Gas Prices Trend Analysis"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

```{r setup}
pacman::p_load(tidyverse, here, scales)
```

## Abstract

Nowadays, the vehicle is one of the important parts in our daily life, as we need to drive vehicles for work, pleasure, school, and many other activities. Consider this, the blood of the vehicles, gas, is sensitive for every American. Before 2018, I rarely paid attention to the gas price. However, the gas price greatly changed and increased almost every day in 2022, during the COVID-19 pandemic. I still remember that I paid around \$7 per gallon for gasoline in California in late 2022. I noticed the gas price can undergo significant changes due to special events, like economic events, pandemic events, and others. I became interested in understanding how gas prices can be changed during special events. It led me to choose Weekly US Gas Prices for my final project. It can answer me the questions that are related to gas price, for example, how a special event, like the COVID-19 pandemic, influences the gas price.

The data is from the U.S. Energy Information Administration (EIA). The variables include "date", "fuel", "grade", "formulation", and "price". The dates are between 1990 and 2025, which is enough for me to analyze the price change for a couple of special events. The fuel is the type of fuel, like gasoline or diesel. The grade is the grade for gas, like regular, mid-grade, or premium. The formulation only applies to gasoline, such as conventional and reformulated. The price is the average US retail price per gallon in US currency. I am able to do analysis for my questions by using this dataset.

# Question 1: How did US Fuel prices change during major events between 2002 and 2025? (the Iraq War 2003, the Global Financial Crisis 2008, and the COVID-19 Pandemic 2020-2022.)

## Introduction

For this question, I will focus on three major events between 2002 and 2025. The three major events are the Iraq War (2023), the 2008 Global Financial Crisis (2008), and the COVID-19 Pandemic (2020). By comparing fuel price trends across these major events, we can understand how different major events, war, the Great Depression, and the pandemic, influenced the US fuel market.

## Approach

There have been three major events for analysis: the Iraq War (2003), the 2008 Global Financial Crisis (2008), and the COVID-19 Pandemic (2020). These events will be shown by the time periods, and highlighted on the visualizations by using shaded regions in the plots.

First of all, I will check if different gasoline grades have different price trends between 2002 and 2025. Then, I will make the annotated time series line chart. It will show the gasoline and diesel prices. If the price trends are similar for different gasoline grades between 2002 and 2025, then I will just use grade "all" for gasoline. The shaded regions are for major event periods. The last step is a faceted line plot. It will show all available fuel grades, which include gasoline and diesel. The separate panels can be used for each fuel grade, and the shaded regions will be used for major event periods.

## Analysis

```{r 1-data}
gas_prices <- read_csv(here("data", "weekly_gas_prices.csv")) |>
  mutate(date = as.Date(date))

glimpse(gas_prices)
```

```{r 1-events}
events <- tribble(
  ~ event_name,
  ~ event_start,
  ~ event_end,
  "The Iraq War",
  as.Date("2003-01-01"),
  as.Date("2003-06-01"),
  
  "The 2008 Global Financial Crisis",
  as.Date("2007-08-09"),
  as.Date("2009-06-30"),
  
  "The COVID-19 Pandemic",
  as.Date("2020-03-11"),
  as.Date("2023-05-05")
)
```

```{r 1-compare-gasoline}
compare_gasoline <- gas_prices |>
  filter(
    fuel == "gasoline",
    grade %in% c("all", "regular", "midgrade", "premium"),
    formulation == "all",
    date >= as.Date("2002-01-01")
  )

ggplot(compare_gasoline, aes(x = date, y = price, color = grade)) +
  geom_line() +
  scale_color_discrete(breaks = c("all", "regular", "midgrade", "premium")) +
  labs(
    title = "Compare Gasoline Price by Grade",
    x = "Year",
    y = "Price per Gallon",
    color = "Gasoline Grade"
  )
```

```{r 1-fuel-compare-data}
fuel_compare <- gas_prices |>
  filter(date >= as.Date("2002-01-01")) |>
  filter(
    (fuel == "gasoline" &
       grade == "all" &
       formulation == "all") |
      (fuel == "diesel" &
         grade == "all")
  )
```

```{r 1-plot, fig.width=15, fig.height=5}
ggplot(fuel_compare, aes(x = date, y = price, color = fuel)) +
  geom_rect(
    data = events,
    aes(
      xmin = event_start,
      xmax = event_end,
      ymin = -Inf,
      ymax = Inf,
      fill = event_name
    ),
    inherit.aes = FALSE,
    alpha = 0.5
  ) +
  geom_line(linewidth = 0.6) +
  scale_x_date(date_breaks = "1 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_color_manual(values = c("gasoline" = "blue", "diesel" = "red")) +
  labs(
    title = "Gasoline and Diesel Prices during Major Events",
    subtitle = "Prices from 2002 to 2025",
    x = "Year",
    y = "Price per gallon",
    color = "Fuel",
    fill = "Major Event"
  ) +
  theme_minimal()
```

```{r 1-fuel-all}
fuel_all <- gas_prices |>
  filter(formulation == "all" | is.na(formulation))
```

```{r 1-all-plot, fig.width=15, fig.height=5}
ggplot(fuel_all, aes(x = date, y = price)) +
  geom_rect(
    data = events,
    aes(
      xmin = event_start,
      xmax = event_end,
      ymin = -Inf,
      ymax = Inf,
      fill = event_name
    ),
    inherit.aes = FALSE,
    alpha = 0.4
  ) +
  geom_line(linewidth = 0.43) +
  facet_wrap(~ fuel + grade) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "U.S. Fuel Prices by Fuel Grade",
    subtitle = "Prices from 2002 to 2025",
    x = "Year",
    y = "Price per gallon",
    fill = "Major event"
  ) +
  theme_minimal()
```

## Discussion

1.  By checking the plot, "Compare Gasoline Price by Grade", I noticed that all the gasoline grades have similar price trends; the only difference is that "premium" and "midgrade" are always higher than "regular", but it is normal, as they are always higher than regular. Therefore, using grade "all", which is the average price for all grades, will be good enough for analysis.

2.  The Iraq War (2003): The war started around March 2003. Due to the war, gas and diesel prices increased by almost 17%, from \$1.5 to \$1.75. It suggests that the oil supply was affected due to the war/ However, the Iraq War ended around May 2003, then the prices immediately returned to the prices that were observed before the war. It states that the war can affect the fuel price in the short term; once the war ends, the price will be back to the normal level.

3.  The 2008 Global Financial Crisis: When the financial crisis occurred, the fuel prices increased. During the middle of the crisis, the fuel prices increased to the highest point, with gasoline increasing by around 33% increased for gasoline (\$4.1 per gallon) and 58% increased for diesel (\$4.75 per gallon) for diesel compared to the price before the crisis occurred. When the financial crisis started in the middle of 2008, the fuel prices crashed for both fuels: reduced 64% for gasoline (\$1.7) and 58% for diesel (\$2). At the end of the financial crisis, the price increased little by little. The financial crisis has a big influence on the fuel price due to the economic market.

4.  The COVID-19 Pandemic: When the pandemic started in 2020, the fuel price did not change too much. However, after a year, when 2021 entered, the fuel price climbed a lot, from around \$2.2 for gasoline and \$2.4 for diesel, to the highest point, around \$5 for gasoline and \$5.75 for diesel in the middle of 2022. Then the price started to reduce. When the pandemic ends in the middle of 2023, the price will be around \$3.75 for both gasoline and diesel. The pandemic brought big chaos to the fuel market.

5.  The plot US Fuel Prices by Fuel Type and Grade shows the price trends and major events for different fuel types and grades.

6.  The trends for gasoline and diesel were almost the same from the plots; however, in some of the highest points, the diesel may have increased more than the gasoline.

# Question 2: How do fuel grade and formulation influence gasoline prices, and how the price differences changed over time?

## Introduction

For this question, I will focus on the prices for all available years. The gasoline has different grades, which include regular, midgrade, and premium. They are also produced by different formulations, conventional and reformulated. This analysis focuses on how fuel characteristics, grades, and formulations influence gasoline prices. We can understand the US gasoline market structure and help customers make a choice when purchasing vehicles, as different vehicles use different fuels.

## Approach

My first step is comparing the gasoline process in different grades (regular, midgrade, and premium). A line chart will be used to show the long-term price for each grade. I can check if different gasoline grades have similar trends or not.

My second step is calculating the price differences between different grades (regular and midgrade, regular and premium, and midgrade and premium). Then, I will make three line charts, and each one will show one price difference.

Then, I will make a box plot. The box plot will show the distributions of gasoline prices across grades. It is easier for me to check the median price and the outlier prices.

My final step is checking the gasoline formulations. I will use conventional and reformulated gasoline to compare each gasoline grade. The price difference between reformulated and conventional gasoline will be calculated and visualized by using faceted line plots.

## Analysis

```{r 2-gas-grade}
gasoline_grade <- gas_prices |>
  filter(
    fuel == "gasoline",
    grade %in% c("regular", "midgrade", "premium"),
    formulation == "all"
  )

ggplot(gasoline_grade, aes(x = date, y = price, color = grade)) +
  geom_line() +
  scale_color_manual(values = c(
    "regular" = "blue",
    "midgrade" = "red",
    "premium" = "green"
  )) +
  scale_color_discrete(breaks = c("regular", "midgrade", "premium")) +
  labs(
    title = "Gasoline Prices by Grade",
    x = "Year",
    y = "Price per Gallon",
    color = "Gasoline Grade"
  ) +
  theme_minimal()
```

```{r 2-price-diff}
price_diff <- gas_prices |>
  filter(fuel == "gasoline",
         grade %in% c("regular", "midgrade", "premium"),
         formulation == "all") |>
  select(date, grade, price) |>
  pivot_wider(names_from = grade, values_from = price) |>
  mutate(diff_reg_mid = midgrade - regular) |>
  mutate(diff_reg_prem = premium - regular) |>
  mutate(diff_mid_prem = premium - midgrade)
```

```{r 2-reg-pre}
ggplot(price_diff, aes(x = date, y = diff_reg_prem)) +
  geom_line(color = "royalblue") +
  labs(title = "Regular and Premium Price Difference", x = "Year", y = "Price Difference") +
  theme_minimal()
```

```{r 2-mid-pre}
ggplot(price_diff, aes(x = date, y = diff_mid_prem)) +
  geom_line(color = "royalblue") +
  labs(title = "Midgrade and Premium Price Difference", x = "Year", y = "Price Difference") +
  theme_minimal()
```

```{r 2-dis-gas-grade}
ggplot(gasoline_grade, aes(x = grade, y = price, fill = grade)) +
  geom_boxplot() +
  labs(title = "Distribution of Gasoline Prices by Grade", x = "Grade", y = "Price per Gallon") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r 2-form-price-diff}
form_price_diff <- gas_prices |>
  filter(
    fuel == "gasoline",
    grade %in% c("regular", "midgrade", "premium"),
    formulation %in% c("conventional", "reformulated")
  ) |>
  select(date, grade, formulation, price) |>
  pivot_wider(names_from = formulation, values_from = price) |>
  mutate(price_difference =
           reformulated - conventional)
```

```{r 2-form-price-diff-plot}
ggplot(form_price_diff, aes(x = date, y = price_difference)) +
  geom_line(color = "darkblue") +
  facet_wrap( ~ factor(grade, levels = c("regular", "midgrade", "premium")), ncol = 1) +
  labs(title = "Price Difference between Reformulated and Conventional Gasoline (Ref. - Con.)", x = "Year", y = "Price Difference") +
  theme_minimal()
```

## Discussion

1.  By checking the Gasoline Prices by Grade, I can see that all of the grades have similar price trends between 1995 and 2025.

2.  By checking the three price difference plots, I can see that the difference between regular gasoline and higher grades of gasoline is increasing year by year, which means the higher grades of gasoline will cost more in the long term. For example, the regular and premium price difference is lower than \$0.25 before 2010, but in 2025, the difference is close to \$0.8, almost one dollar. The vehicles that will use higher grades of gasoline will cost the driver more in the long term.

3.  The Distribution of Gasoline Prices by Grade shows the median price for each grade. The premium grade has the highest outlier compared with the others.

4.  The Price Difference between Reformulated and Conventional Gasoline shows that reformulated gasoline is always more expensive than conventional. The difference increased over time between reformulated and conventional gasoline. There had been a significant increase in reformulated and conventional gasoline after 2020.
