---
title: "Final Project - US GAS Prices Analysis"
subtitle: "Proposal"
author: 
  - name: "Zachary Zhicheng Xu"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "US Gas Prices Analysis"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(here)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

# Remote reading
weekly_gas_prices_remote <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-07-01/weekly_gas_prices.csv')

# Local reading
weekly_gas_prices_local <- read_csv(here("data", "weekly_gas_prices.csv"))

glimpse(weekly_gas_prices_local)

```

Nowadays, the vehicle is one of the important parts in our daily life, as we need to drive vehicles for work, pleasure, school, and many other activities. Consider this, the blood of the vehicles, gas, is sensitive for every American. Before 2018, I rarely paid attention to the gas price. However, the gas price greatly changed and increased almost every day in 2022, during the COVID-19 pandemic. I still remember that I paid around \$7 per gallon for gasoline in California in late 2022. I noticed the gas price can undergo significant changes due to special events, like economic events, pandemic events, and others. I became interested in understanding how gas prices can be changed during special events. It led me to choose Weekly US Gas Prices for my final project. It can answer me the questions that are related to gas price, for example, how a special event, like the COVID-19 pandemic, influences the gas price.

The data is from the U.S. Energy Information Administration (EIA). The variables include "date", "fuel", "grade", "formulation", and "price". The dates are between 1990 and 2025, which is enough for me to analyze the price change for a couple of special events. The fuel is the type of fuel, like gasoline or diesel. The grade is the grade for gas, like regular, mid-grade, or premium. The formulation only applies to gasoline, such as conventional and reformulated. The price is the average US retail price per gallon in US currency. I am able to do analysis for my questions by using this dataset.

## Questions

1.  How did US Fuel prices change during major events between 1990 and 2025? (the Iraq War 2003, the Global Financial Crisis 2008, and the COVID-19 Pandemic 2020-2022.)
2.  How do fuel grade and formulation influence gasoline prices, and how the price differences changed over time?

## Analysis plan

- First question:

  - Variables may use:

    - date
    - fuel
    - grade
    - price

  - Variables may create:

    - event_name \*
    - event_start
    - event_end
    - event_type

  - Visualizations:

    - Annotated Line plots
      - Long-term gas price trends
      - Highlight major event periods using shaded regions and labels
    - Faceted line plots
      - Compare responses across fuel categories

    (\* Additional visualizations may be developed depending on the analysis progress.)

- Second question:

  - Variables may use:

    - date
    - grade
    - formulation
    - price

  - Variables may create:

    - year (from date)
    - average_price (grade and formulation)
    - price_difference (Premium - Regular)

  - Visualizations:

    - Line plots

      - Long-term trends by grade
      - Changes in price gaps over time

    - Box plots

      - Compare price distributions by grades
      - identify variability and outlier

    (\* Additional visualizations may be developed depending on the analysis progress.)

- Narrative

  - All visualizations will be accompanied by narrative explanations (key trends, event impacts, differences among fuel types, grades, and formulations)
